---
title: "Help using the 8-Year Index Annuity Calculator /w Biennial S&P500 Indexing"
output: html_document
---

Unfortunately, this app will make little sense without some domain knowledge.  This helpfile is meant to guide you to an understand of exactly what I am calculating.


## Annuity?

An annuity a specific type of life insurance policy.  For this app, we are specifically concerned with "Single Premium Lump Sum" annuities with an 8-year term.  This means, if you are buying a $10000 annuity, at the start of the policy you pay the life insurance company $10000.  At the end of 8 years, you will get a check for some amount, and the policy is over.

In terms of financial instruments you are familiar with, this is much closer to a CD or Savings Bond than what you traditionally think of as an Annuity.  Why you would want a Single Premium Lump Sum annuity over these instruments is complex -- but it basically comes down to how the gains are taxed.

## Fixed Annuity?

Traditionally, the annuity we describe pays a fixed interest rate, for example, 2%.  This means that after 8 years, your $10000 "Simple Premium Lump Sum" annuity would be worth:

```{r}
10000 * 1.02^8
```

## The problem with fixed annuities

The problem with fixed annuities that we are trying to solve is the interest rate.  In short, most fixed annuities pay a very low interest rate.  For example, today, a 10-year single premium lump sum annuity would pay around 2.8%.  This is favorable compared to CDs and Government Bonds, especially after tax concerns are taken into account, but compared to corporate bonds and the stock market it's pathetic.

CDs and Money Market accounts really are the most apt comparison because if you're buying an annuity they are structured s.t. it's impossible to lose money.  Still though, considering the S&P500 annual return over the last 30 years is somewhere around 8% -- we might want a little exposure assuming we could somehow keep intact the guarantee we cannot lose money.

## The solution -- Indexed Annuities

The solution to this is indexed annuities.  Indexed annuities have a variable return based on some market, however that return is guaranteed to be above 0%.  There are tons of different types of indexed annuities out there.  The problem is the formulas they use to calculate returns is complex, and most firms (I'm looking at you Metlife and JNL!) don't provide detailed historical returns.

## The purpose of this Application

This application is attempting to provide detailed historical returns for one very popular type of index annuity.  You provide 3 policy parameters plus a policy start date range, and the application will calculate the equivalent fixed annuity interest rate if you purchased a policy on each date in that range.

## Input Variables and Calculated Return

The input variables are:

* *Guaranteed Minimum Interest Rate*: At the end of 8 years, the policy is guaranteed to pay out, at *minimum*, the equivalent of a fixed interest policy with this interest rate.
* *Initial Premium Bonus Percentage*: To entice you to buy this product, the life insurance policy will credit the policy bonus cash at the very beginning of the policy.  For example, if the bonus rate is 0.05 (5%), on a $10,000 policy you will start with an initial value of $10,500.  Note the GMIR applies to this value.
* *Policy Start Dates*: This is the range of dates we will calculate the equivalent fixed interest rate over.  For example, if the range is '1999-12-31' to '2000-12-31', we will calculate equivalent fixed annuity interest rate for every day in the year 2000.  We want a range of dates to get a sense of the distribution of possible returns.
* *Biennial Interest Rate Cap*: This is the big one.  The life insurance company has a cap on how much of the returns of the stock market they will credit to your account.  This interest rate is is that cap.  For an eight year policy, we have 4 2-year periods.  At the end of each 2-year period, we will look at the return of the S&P500 for that period.  If it is positive, we will credit either the 2-year return (not including dividends) or the Biennial Interest Rate Cap -- whichever is lower.  If the return is negative, we will not deduct any money -- essentially we are crediting $0.

### The calculation

   The easiest way to show you how all of these values turn into a real number is to calculate what our return would have been for some inputs.  Let us assume:
   
   * We bought the policy on January 3rd, 2000
   * The GMIR is 0.55%
   * The initial premium bonus is 5%
   * The Biennial Interest Rate Cap is 8%
   * The policy is for $10,000

As a small side-note, if you were to buy this policy today, those numbers are fairly realistic.

First, we are going to calculate what our guaranteed minimum return is using the bonus and the GMIR.

```{r}
10000*1.05*(1.0055)^8
```

So no matter what, we will get back at least $10,970.99 after 8 years.  Considering we had the option to buy a fixed interest policy would have paid 2.5% for a value of $12,184.03 we might be a little upset.  That's the downside to these policies -- the worst case scenario is always worse than a fixed interest policy.

To calculate how much we actually would have gotten back, we need the S&P500 closing prices for the date range.  If it is a weekend or holiday, we will use the close on the previous Friday.  They are as follows:

* 2000-01-03 : 1455.22
* 2002-01-03 : 1165.27
* 2004-01-03 : 1108.48
* 2006-01-03 : 1268.80
* 2008-01-03 : 1447.16

We can then calculate the percentage gain/loss:

* 2000-01-03 to 2002-01-03 : -24.88%
* 2002-01-03 to 2004-01-03 : -5.12%
* 2004-01-03 to 2006-01-03 : +12.64%
* 2006-01-03 to 2008-01-03 : +12.32%

We hit the cap twice, and the market had negative returns twice.  Based on this, the calculated return would be:

```{r}
initialAmount <- 10000;
bonusPercentage <- 0.05;
rateCap <- 0.08;
spxRates <- c(-0.2488, -0.0512, 0.1264, 0.1232);

#The minimum return over a period is 0%
spxRates <- sapply(spxRates, function(x) { max(0, x)});
#The maximum return over a period is the cap.
spxRates <- sapply(spxRates, function(x) { min(rateCap, x)});

total <- initialAmount * (1+bonusPercentage) * prod(1+spxRates);
total

```

Or if seeing it written out makes it more clear:

```{r}
10000 * (1 + 0.05) * (1+0) * (1+0) * (1+0.08) * (1+0.08)
```

This is higher than the GMIR, so this is the amount of money we actually get back.  To convert it to a fixed interest rate over 8 years, we take the 8th root of the relative gains and subtract 1.

```{r}
equivalentFixedInterestRate <- ((total/initialAmount)^(1/8))-1
equivalentFixedInterestRate
```

And just to check our work, we can calculate the return of a fixed index policy with that rate:

```{r}
10000*(1+equivalentFixedInterestRate)^8
```

Even though we hit two really rough periods, we still did better than a 2.5% fixed rate policy.  We did get a bit lucky though -- consider (or use the app!) what would have happened if we bought the policy several months later -- hitting the worst of the 2000 dot-com crash, and the 2007 housing crisis (which bottomed out sometime much later in 2008).

## Some Notes

   First, if you're wondering, the app does return the same value we calculated above.
   
   Second, in the original version I have R code (via quantmod) that hits Yahoo for S&P data.  In this version though, I am using a local file just in case something happens during the grading period, because of this -- the max value for the Policy Start Date is 11/23/2006 (corresponding to a policy that pays out on 11/23/2014).
   
   Third, the min value of "Policy Start Date" is 1970-01-01 -- as far back as Yahoo goes.  There is some weirdness in the Date Range shiny control where it looks like dates before the first year in the decade are inaccessible.
   
   Fourth, sorry for the calculation times.  I spent some time optimizing code, but in the end this is the best I could do in the time I had without bringing in multithreading libraries.  There is a progress bar at the top so you can at least see how far along it is.